Assessing the impacts of time to detection distribution assumptions on detection probability estimation

### 2017-04-06 Draft.tex
Updates to manuscript.

@@

\newcommand{\pdet}{p^{(det)}}
\newcommand{\ind}{\stackrel{ind}{\sim}}
\newcommand{\Fm}{F_T^{(M)}}
\newcommand{\nobs}{n^{(obs)}}

@@

Abundance estimates from animal point-count surveys require accurate estimates of detection probabilities.  
Removal sampling, where individuals are solely counted on their first capture, provides one established methodology for estimating detection probabilities \citep{Farnsworth2002}.
% cost-effective in that it requires neither multiple observers nor multiple site visits.
%Removal sampling is a species abundance surveying methodology in which observers capture and physically remove animals at a survey site over a series of trapping sessions.
%Assuming that individual capture rates remain constant over all trapping sessions, it is possible to estimate the proportion of individuals still not captured from the pattern of captures over time .  
%\citet{Farnsworth2002} adapted removal sampling to point-count surveys, suggesting that if: (i) observers record only the first time-to-detection for each animal, and (ii) times to first detection are interval-censored, then data will be similar in for to traditional removal sampled data.
%In the analysis of such data, the observation period is subdivided into unit intervals of equal duration, each of which is equivalent to a trapping session.
%Farnsworth's removal assumptions:\\
%(i) Closed population during survey\\
%(ii) no double-counting\\
%(iii) easy-to-detect group is entirely counted during first interval\\
%(iv) TTDD for hard-to-detect group is correct after the end of the observation period\\
%(v) if using limited radius counts, they are accurate
% Johnson2008: all birds present are available for detection
A typical assumption in removal sampling is a constant detection rate throughout the observation period, but this assumption is often unjustified \citep{Alldredge2007, LeeMarsden2008}.  
In particular, animal behaviors such as intermittent singing in birds and frogs or diving in whales \citep{Scott2005, Diefenbach2007, Reidy2011}, differences in behavior across subgroups of animals \citep{Otis1978, Farnsworth2002}, observer impacts on animal behaviors \citep{McSheaRappole1997, Rosenstock2002, Alldredge2007}, and variations in observer effort, e.g. saturation or lack of settling down period \citep{Petit1995, LeeMarsden2008, Johnson2008}, can all lead to time-varying rates of detection.

@@

Unlike most survival analyses, the number of individuals $N$ present at a survey is unknown and may be the primary quantity of interest.  
We embed the TTDD in a hierarchical framework for multinomial counts called an N-mixture model \citep{Wyatt2002, Royle2004NMixture}, which is an entirely different use of `mixture' from the mixture models in the previous paragraph.
For our purposes, the N-mixture framework provides three clear benefits: 1) it handles counts within a flexible multinomial data framework \citep{RoyleDorazio2006} which accords with the interval-censored data collection that is customary in point-count surveys \citep{Ralph1995}, 2) the hierarchical structure readily lends itself to including abundance- and detection-related covariates and random effects \citep{Dorazio2005, Etterson2009, Amundson2014}, and 3) for a Bayesian analysis, we can sample the posterior joint distribution of N-mixture parameters straight-forwardly using Markov chain Monte Carlo (MCMC).  
The N-mixture framework models abundance as a latent variable with a Poisson or other discrete distribution and independently models detection probabilities.  
Several previous studies have employed the N-mixture framework to analyze removal sampled point-count data  \citep{Royle2004Generalized, Dorazio2005, Etterson2009, Solymos2013, Amundson2014, Reidy2016}.

@@

Framing a model in terms of time-to-detection leads to two practical differences vis-a-vis constant-detection models.  
First, in order to model covariate and random effects on detection, we perform mixed effects linear regression on the log of the rate parameter as in \citet{Solymos2013}, whereas most existing studies instead construct regression models on the logit of the equal-interval detection probability.  
The latter is not possible when detection rates are not constant.  
Second, because we can obtain interval-specific detection probabilities from the TTDD by partitioning its cdf, we can directly model the data according to their existing interval structure rather than subdividing the observation period into intervals of equal duration.  
Indeed our model fits exact time-to-detection data, whereas existing constant-detection removal models only approximate exact data by subdividing the observation interval into a large number of fine equal-duration intervals \citep{Reidy2011, Amundson2014}.

@@

\begin{figure}\centering
\includegraphics[width=\textwidth]{"Oral Prelim/AMRO_JABES"}
\caption{\label{fig:schematic} 
Illustration of fitting mixture exponential (left) and mixture gamma (right) time-to-detection distributions (TTDDs) to multinomial removal-sampled observations (center) assuming no survey-level heterogeneity.
The mixture TTDD consists of a continuous TTDD (thick line) plus a mixture component (light gray rectangle), which constitute $\gamma$ and $1-\gamma$ proportions of the population, respectively.
We estimate $\pdet$ as the proportion of the TTDD before the end of the observation period $C$, leaving an estimated proportion $1-\pdet$ undetected.
}
\end{figure}

@@

\adam{I think we start by describing our joint likelihood: p(counts $| N, \pdet$) and p(times $|$ counts,TTDD) and prior.
We explain that, in isolation, $N$ and $\pdet$ are not identifiable in the first piece, which is why we use the second piece to inform our estimates of $\pdet$.
Then, in the body of section \ref{sec:exact_time}, we introduce a second paragraph which defines our distribution for $\nobs$ and completes the data likelihood.
The discussion of identifiability would travel with the distribution of $\nobs$ and be cleaned up a little.
Perhaps even split the discussion of TTDDs into another subsection.
}

@@

\subsection{Exact time to detection}\label{sec:exact_time}

Suppose that, for each survey $s$ ($s=1,\dots,S$), $N_s$ individuals are present.  
Imagine an observer could remain at the survey location until every individual is detected, recording the time to detection $t_{sb}$ (for bird, $b=1,\dots,N_s$) for each.
Assuming detection times for all individuals at a survey are independent, identically distributed according to a common time-to-detection distribution (TTDD), we define $T_{sb}$ as a random variable with cumulative distribution function (cdf) $F_T(t)$ and probability density function (pdf) $f_T(t)$.
In reality, times to first detection are often truncated due to a finite survey length of $C$, meaning that each individual has a detection probability $\pdet=F_T(C)$.  
The conditional distribution of observed detection times consequently has pdf $f_{T|det}(t)= f_T(t)/F_T(C)$ for $0<t<C$, cdf $F_{T|det}(t) = \int_0^t f_{T|det}(x) dx$, and instantaneous detection rate, or hazard function, $h(t) = f_T(t) / [1-F_T(t)]$. 
%$T_{sb|det} \ind f_{T|det}(t_{sb})$ where $T_{sb|det}$ is observed only for those $n_s^{(obs)}$ individuals whose observed time to first detection is less than $C$, i.e. $t_{sb}<C$.

A common choice for TTDD is an exponential distribution, i.e. $T_{sb}\ind \mbox{Exp}(\varphi)$, which imposes a constant first detection rate, i.e. $h(t) = \varphi$.
Choosing another TTDD can allow for a systematic non-constant detection regime. 
For example, to model an observer effect where: (i) the observer's arrival suppresses or stimulates detectable cues, but (ii) organisms acclimate and gradually return to constant detection, a gamma TTDD could be appropriate.
%In addition to an exponential and gamma TTDD, we also consider Weibull and lognormal as these distributions are often used in survival analysis. 
Like the gamma TTDD, Weibull and lognormal TTDDs offer the flexibility of a two-parameter form and allow rates to either increase or decrease during the survey; however, the shapes of these distributions differ, most notably in tail probability.
All three may be reasonable TTDD choices if, for instance, detection rates vary among individuals.
This leads to marginal detection rates that decrease over time as individuals with higher detection rates are removed from the population.

To facilitate the later inclusion of fixed and random effects, we use the following rate-based parameterizations: $T\sim \Exp(\varphi), E[T]=1/\varphi$; $T\sim \Ga(\alpha,\varphi), E[T] = \alpha/\varphi$; $T\sim \We(\alpha,\varphi), E[T]=\mathrm{\Gamma}(1+1/\alpha)/\varphi$; and $T\sim \LN(\varphi,\alpha), E[T] = \exp(\alpha^2/2)/\varphi$.  
This parameterization of the lognormal relates to the standard ($\mu, \sigma^2$) parameterization by $\varphi = \exp(-\mu)$ and $\alpha = \sigma$.  %$\mu = -\log(\varphi)$
The exponential distribution is a special case of both the gamma and Weibull distributions when $\alpha=1$. 
We employ a log link for $\varphi$, and therefore our model is equivalent to a generalized linear model with a log link on the mean detection time.

We model the number of individuals at survey $s$ for which $t_{sb}$ is actually observed as $n_{s}^{(obs)} \ind \mbox{Binomial}\left(N_{s}, \pdet\right)$. 
Using maximum likelihood, we can estimate the parameters in a TTDD from exact times to first detection and thus estimate $\pdet$ and its uncertainty.
With an estimate and uncertainty for $\pdet$, we are in a scenario of a binomial model with unknown $N_s$ and ``known'' $p$ and thus can estimate $N_s$.
In these models, point estimates of $N_s$ are unstable unless $\pdet>0.41$ \citep{Olkin1981}.
One approach to regularizing these estimates is to construct a hierarchical model for the site-specific abundance, e.g. $N_s\ind \Po(\lambda)$ \citep{Raftery1988, Royle2004NMixture}. 
With this assumption, we can decompose $N_{s}$ into observed and unobserved portions: $n^{(obs)} \sim Po(\lambda \pdet)$ and, independently, $n_{s}^{(unobs)} \sim \Po\left(\lambda[1-\pdet]\right)$.
Although alternative distributions could be considered, e.g. negative binomial, our experience with Ovenbird point counts suggests that, after accounting for appropriate explanatory variables, the resulting abundances are likely underdispersed rather than overdispersed, and thus we will use the Poisson assumption here.

@@

Due to the harried process of avian point counts, times to first detection are typically not recorded exactly, but are instead censored into $I$ intervals. 
Let $C_i$ for $i=1,\dots,I$ indicate the right endpoint of the $i$th interval then $C_I$ is the total survey duration and, letting $C_0=0$, the $i$th interval is $(C_{i-1},C_{i}]$. 
Let $n_{si}$ be the number of individuals counted during interval $i$ on survey $s$, $n_{s}^{(obs)} = \sum_{i=1}^I n_{si}$, and $\vn_{s}=(n_{s1},\dots,n_{sI})$.
Assuming independence amongst individuals and sites, we have $\vn_{s} \ind \Mult \left(n_{s}^{(obs)}, \vp_{s}\right)$, where $\vp_{s}=(p_{s1},\dots,p_{sI})$ and is calculated from the TTDD: $p_{si} = F_{T|det}(C_i) - F_{T|det}(C_{i-1})$.

@@

For ease of reference, the final full model is provided in equation \eqref{eq:model} where the conditioning of the TTDD cdf on $\alpha$ and $\varphi_s$ is made explicit.
\begin{align}
n_s^{(obs)} &\ind \Po(\lambda_s \pdet_s) \nonumber\\
\vn_s &\ind \Mult(n_s^{(obs)}, \vp_s); \qquad \vp_s = (p_{s1},\dots,p_{sI}) \nonumber\\
\pdet_s &= F_T(C_I|\alpha,\varphi_s) \nonumber \\
p_{s1} &= \left[(1-\gamma) + \gamma\Fm(C_1|\alpha,\varphi_s)\right]/\pdet_s \label{eq:model} \\
p_{si} &= \gamma\left[\Fm(C_i|\alpha,\varphi_s) - \Fm(C_{i-1}|\alpha,\varphi_s)\right]/\pdet_s \nonumber\\
\log(\lambda_s) &= \vX_{s}^A\vbeta^A + \vZ_{s}^A\vxi^A; \qquad \xi_j^A \ind N(0,\sigma_{A[j]}^2) \nonumber\\
\log(\varphi_{s}) &= \vX_{s}^D\vbeta^D + \vZ_{s}^D\vxi^D; \qquad \xi_j^D \ind N(0,\sigma_{D[j]}^2) \nonumber
\end{align}
We adopt a Bayesian approach and therefore require a prior over the model parameters.
To ease construction of a default prior for this model, we standardize all explanatory variables and then construct priors to be diffuse within a reasonable range of values.
Normal prior mean and standard deviation (sd) for the abundance intercept was set at a median abundance of 3 birds per site and a 95\% probability of 0-14 birds present (counted and uncounted).  
Normal prior mean and sd for the detection intercept were chosen so that, based on an intercept-only non-mixture model with $\alpha=1$: (i) median prior detection probability was $p_{s}^{(det)} = 0.50$, and (ii) 95\% of the prior detection probability was within $p_{s}^{(det)} \in (0.01, 1.0)$.  
Normal priors for fixed effect parameters were centered at zero with standard deviations matching the appropriate intercept term.  
All standard deviations and $\alpha$ were given half-Cauchy priors with location 0 and scale 1 for the untruncated Cauchy, and the mixture parameter $\gamma$ was assigned a Unif(0,1) prior in mixture models.
All scalar parameters were assumed independent \emph{a priori}.

We fit the models by MCMC sampling using the Bayesian statistical software Stan, implemented via the R package \texttt{rstan} version 2.8.0 \citep{Rstan2015}.  
%Stan model code and \texttt{rstan} code to run the models can be found in the Supplementary Materials.  
We discarded half of the iterations as warmup and thinned by 10.  
We monitored convergence of the MCMC chains using Geweke z-score diagnostics \citep{Geweke1991} and reran models if lack of convergence was indicated by a non-normal distribution of the z-scores or if the effective sample size for any parameter was below 1000.  
The number of iterations used depended on the model and is detailed later. 
For most models, we accepted Stan defaults for initial values; however, gamma and Weibull models sometimes failed to run unless care was taken in the specification of initial values.

@@

We focus inference on the overall detection probability across all surveys: $\pdet = \sum\limits_{s}n_{s}^{(obs)}\big/$ $\sum\limits_{s}N_{s}$.  
Within simulation studies, we compared model estimates to true $\pdet$ values by percent coverage and by posterior p-values --- the proportion of MCMC samples having a smaller value than the true value.

@@

In the following analyses we distinguish two categories of purely continuous TTDDs: peaked and nonpeaked.  
Detection rates $h(t)$ of peaked distributions generally increase over time, while detection rates of nonpeaked distributions generally decrease over time.
More formally, we define a peaked TTDD as having a mode greater than zero (or $C_1$ for lognormal) while a non-peaked TTDD has a mode of zero (or less than $C_1$), but we consider exponential TTDDs to be neither peaked nor nonpeaked.

@@

To assess the need for incorporating a mixture component to increase the probability of detection in the initial interval as discussed in Section \ref{s:interval}, we simulated 16 intercept-only datasets from each of 14 TTDDs: each combination of peaked/non-peaked, mixture/non-mixture, and exponential/gamma/Weibull/lognormal, where all exponential models are non-peaked.  
We chose the number of surveys (381) and true parameter values (Table \ref{S-simvals}) to mimic values from the Ovenbird analysis (Section \ref{sec:ovenbirds}).  
In particular, we set parameters such that (i) the overall expected detection probability was 0.80, (ii) for mixture datasets, $\gamma = 0.65$ (meaning 35\% of individuals were immediately detected), (iii) in nonpeaked models, 70\% of \textit{detected} individuals were observed during the first two minutes, and (iv) in peaked models, the detection mode for `hard to detect' individuals occured at 5 minutes.

Each dataset was fit with two models: mixture and non-mixture version of the distribution family, e.g. exponential, used to simulate the data.
For each dataset-analysis combination, we ran 60,000 iterations which showed no evidence of lack of convergence according to the Geweke diagnostic and reached over 1,000 effective samples for all parameters.

We summarized the analysis by calculating the average, across the simulations, of the posterior median of $\pdet$ (Med. $p$), the average proportion of the posterior distribution of $\pdet$ larger than the truth (Q($p$)), and average coverage for 50\% and 90\% credible intervals. If the analyses are providing a reasonable estimate of $\pdet$, we would expect Med. $p$ to be around the true value of 0.8, Q($p$) to be around 0.5 indicating that half of the posterior is above and below the true value of 0.8, and the coverage to be close to their credibility. Table \ref{tbl:sim1} provides a summary of these quantities. 
When a mixture model is used to simulate the data (bottom half of the table), there is clearly a benefit to using a mixture model for inference.  
Using a non-mixture model for inference, the credible interval coverage is near zero for most models with the exponential model overestimating $\pdet$ (Q($p$) $\approx 1$) and the other models underestimating. 
When a non-mixture model is used to simulate the data (top half of the table), there are no clearly discernable differences between the ability of a non-mixture or mixture model to capture $\pdet$. 
For nonpeaked non-mixture data (top 4 lines), the mixture model is less biased for lognormal and Weibull scenarios (Q($p$)$\approx 0.5$), and CIs are 15-40\% narrower (results not shown).
These results support the general default use of a mixture model over a non-mixture model.

\begin{figure}\centering
\includegraphics[width=\textwidth]{"Sims/SimZero/mixture_fig"}
\caption{
Comparisons of data and models with and without mixture TTDDs.  \textit{Left}: Average posterior median detection probabilities across 100 replicate simulations.  Dashed lines show unbiased estimation.   \textit{Right}: Coverage of 50\% credible intervals across 100 replicate simulations.  Dashed lines depict a 95\% range of observed coverage that is consistent with nominal coverage.  \textit{Rows}: Data simulated from non-mixture (upper) or mixture (lower) TTDDs.  \textit{Columns}:  Data simulated from nonpeaked (left), exponential (center), or peaked (right) TTDDs.  In these simulations, each inference model was only fit to datasets from the same TTDD family (e.g. lognormal to lognormal).
}
\label{fig:mixture_fig} 
\end{figure}

>>>>

\renewcommand{\floatpagefraction}{0.8}  % Changes settings for when a figure hogs a page all to itself

@@

\newcommand{\pdet}{p^{(det)}}
\newcommand{\ind}{\stackrel{ind}{\sim}}
\newcommand{\Fm}{F_T^{(M)}}
\newcommand{\dt}{\text{det}}
\newcommand{\ftdt}{f_{T|\dt}(t|\dt)}
\newcommand{\nobs}{n^{(obs)}}

@@

Abundance estimates from animal point-count surveys require accurate estimates of detection probabilities.  
Removal sampling, where individuals are solely counted on their first capture, provides one established methodology for estimating detection probabilities \citep{Farnsworth2002}.
% cost-effective in that it requires neither multiple observers nor multiple site visits.
%Removal sampling is a species abundance surveying methodology in which observers capture and physically remove animals at a survey site over a series of trapping sessions.
%Assuming that individual capture rates remain constant over all trapping sessions, it is possible to estimate the proportion of individuals still not captured from the pattern of captures over time .  
%\citet{Farnsworth2002} adapted removal sampling to point-count surveys, suggesting that if: (i) observers record only the first time-to-detection for each animal, and (ii) times to first detection are interval-censored, then data will be similar in for to traditional removal sampled data.
%In the analysis of such data, the observation period is subdivided into unit intervals of equal duration, each of which is equivalent to a trapping session.
%Farnsworth's removal assumptions:\\
%(i) Closed population during survey\\
%(ii) no double-counting\\
%(iii) easy-to-detect group is entirely counted during first interval\\
%(iv) TTDD for hard-to-detect group is correct after the end of the observation period\\
%(v) if using limited radius counts, they are accurate
% Johnson2008: all birds present are available for detection
A standard assumption in removal sampling is a constant detection rate throughout the observation period, but this assumption is often unjustified \citep{Alldredge2007, LeeMarsden2008}.  
In particular, animal behaviors such as intermittent singing in birds and frogs or diving in whales \citep{Scott2005, Diefenbach2007, Reidy2011}, differences in behavior across subgroups of animals \citep{Otis1978, Farnsworth2002}, observer impacts on animal behaviors \citep{McSheaRappole1997, Rosenstock2002, Alldredge2007}, and variations in observer effort, e.g. saturation or lack of settling down period \citep{Petit1995, LeeMarsden2008, Johnson2008}, can all lead to time-varying rates of detection.

@@

Unlike most survival analyses, the number of individuals $N$ present at a survey is unknown and may be the primary quantity of interest.  
We embed the TTDD in a hierarchical framework for multinomial counts called an N-mixture model \citep{Wyatt2002, Royle2004NMixture} --- which is an entirely different use of `mixture' from the mixture models in the previous paragraph.
For our purposes, the N-mixture framework provides three clear benefits: 1) it handles counts within a flexible multinomial data framework \citep{RoyleDorazio2006} which accords with the interval-censored data collection that is customary in point-count surveys \citep{Ralph1995}, 2) the hierarchical structure readily lends itself to including abundance- and detection-related covariates and random effects \citep{Dorazio2005, Etterson2009, Amundson2014}, and 3) for a Bayesian analysis, we can sample the posterior joint distribution of N-mixture parameters straight-forwardly using Markov chain Monte Carlo (MCMC).  
The N-mixture framework models abundance as a latent variable with a Poisson or other discrete distribution and independently models detection probabilities.  
Several previous studies have employed the N-mixture framework to analyze removal sampled point-count data  \citep{Royle2004Generalized, Dorazio2005, Etterson2009, Solymos2013, Amundson2014, Reidy2016}.

@@

Framing a model in terms of time-to-detection leads to two practical differences vis-a-vis constant-detection models.  
First, in order to model covariate and random effects on detection, we perform mixed effects linear regression on the log of the rate parameter as in \citet{Solymos2013}, whereas most existing studies instead construct regression models on the logit of the equal-interval detection probability.  
The latter is not possible when detection rates are not constant.  
Second, because we can obtain interval-specific detection probabilities from the TTDD by partitioning its cdf (Figure \ref{fig:schematic}), we can directly model the data according to their existing interval structure rather than subdividing the observation period into intervals of equal duration.  
Indeed our model fits exact time-to-detection data, whereas existing constant-detection removal models only approximate exact data by subdividing the observation interval into a large number of fine equal-duration intervals \citep{Reidy2011, Amundson2014}.

@@

\begin{figure}\centering
\includegraphics[width=\textwidth]{"Oral Prelim/AMRO_JABES"}
\caption{\label{fig:schematic} 
Illustration of fitting mixture exponential (left) and mixture gamma (right) time-to-detection distributions (TTDDs) to interval-censored removal-sampled observations (center).
The mixture TTDD consists of a continuous TTDD (thick line) plus a mixture component of first-interval detections (light gray rectangle), constituting $\gamma$ and $1-\gamma$ proportions of the population, respectively.
We estimate $\pdet$ as the proportion of the TTDD before the end of the observation period $C$, leaving an estimated proportion $1-\pdet$ undetected.
}
\end{figure}

@@

@@

\subsection{Distributions for exact times to detection}\label{sec:exact_time}

Suppose that, for each survey $s$ ($s=1,\dots,S$), $N_s$ individuals are present.  
Imagine an observer could remain at the survey location until every individual is detected, recording the time to detection $t_{sb}$ (for bird, $b=1,\dots,N_s$) for each.
Assuming detection times for all individuals at a survey are independent, identically distributed according to a common time-to-detection distribution (TTDD), we define $T_{sb}$ as a random variable with cumulative distribution function (cdf) $F_T(t)$ and probability density function (pdf) $f_T(t)$.
In practice, times to first detection are truncated due to a finite survey length of $C$, meaning each individual has detection probability $\pdet=F_T(C)$.  
The conditional distribution of observed detection times consequently has pdf $\ftdt= f_T(t)/F_T(C)$ for $0<t<C$, cdf $F_{T|\dt}(t|\dt) = \int_0^t f_{T|\dt}(x|\dt) dx$, and instantaneous detection rate, or hazard function, $h(t) = f_T(t) / [1-F_T(t)]$. 
%$T_{sb|\dt} \ind f_{T|\dt}(t_{sb}|\dt)$ where $T_{sb|\dt}$ is observed only for those $\nobs_s$ individuals whose observed time to first detection is less than $C$, i.e. $t_{sb}<C$.

A common choice for TTDD is an exponential distribution, i.e. $T_{sb}\ind \mbox{Exp}(\varphi)$, which imposes a constant first detection rate, i.e. $h(t) = \varphi$.
By choosing another TTDD, we can allow for a systematic non-constant detection regime. 
For example, to model an observer effect where: (i) the observer's arrival suppresses or stimulates detectable cues, but (ii) individuals acclimate and gradually return to constant detection, a gamma TTDD would be appropriate.
Like the gamma TTDD, Weibull and lognormal TTDDs offer the flexibility of a two-parameter form and allow rates to increase or decrease during the survey.
All three TTDDs may provide reasonable empirical approximations of non-constant detection, though the shapes of the distributions differ, potentially leading to differing inference.
For instance, when detection rates vary across individuals, the result is a marginal detection rate that decreases over time.
Whether the marginal rate is best approximated by a gamma, lognormal, Weibull, or some other TTDD depends on just how rates vary across individuals.

To facilitate the later inclusion of fixed and random effects, we use the following rate-based parameterizations: $T\sim \Exp(\varphi), E[T]=1/\varphi$; $T\sim \Ga(\alpha,\varphi), E[T] = \alpha/\varphi$; $T\sim \We(\alpha,\varphi), E[T]=\mathrm{\Gamma}(1+1/\alpha)/\varphi$; and $T\sim \LN(\varphi,\alpha), E[T] = \exp(\alpha^2/2)/\varphi$.  
This parameterization of the lognormal relates to the standard ($\mu, \sigma^2$) parameterization by $\varphi = \exp(-\mu)$ and $\alpha = \sigma$.  %$\mu = -\log(\varphi)$
The exponential distribution is a special case of both the gamma and Weibull distributions when $\alpha=1$. 
We employ a log link to model $\varphi$, and therefore our model is equivalent to a generalized linear model with a log link on the mean detection time.

\subsection{TTDDs in an N-mixture model}

A basic N-mixture model describes observed survey-level abundance $\nobs$ with a hierarchy where $\nobs_s \ind \mbox{Binomial}\left(N_{s}, \pdet\right)$ and $N_s \ind Po(\lambda)$.
We can decompose $N_{s}$ into observed and unobserved portions: $\nobs \ind Po(\lambda \pdet)$ and, independently, $n_{s}^{(unobs)} \ind \Po\left(\lambda[1-\pdet]\right)$.
Although alternative distributions can be considered, e.g. negative binomial, our experience with Ovenbird point counts suggests that, after accounting for appropriate explanatory variables, the resulting abundances are likely underdispersed rather than overdispersed, and thus we will use the Poisson assumption here.

The above definitions complete our exact-time homogenous-survey data model, consisting of distributions for counts and observed detection times:
\begin{align}
\nobs_s &\ind \Po(\lambda \pdet) \nonumber\\
p(\textbf{t}_s) &= \prod\limits_{b=1 \dotso\nobs_s} f_{T|\dt}(t_{sb} | \det, \alpha, \varphi) \label{eq:simplemodel}\\
\pdet &= F_T(C|\alpha,\varphi) \nonumber
\end{align}
where $\textbf{t}_s$ is a vector of observed times at survey $s$.

%We model the number of individuals at survey $s$ for which $t_{sb}$ is actually observed as $\nobs_s \ind \mbox{Binomial}\left(N_{s}, \pdet\right)$. 
%Using maximum likelihood, we can estimate the parameters of a TTDD from the observed times to first detection and thereby estimate $\pdet$ and its uncertainty.
%With an estimate and uncertainty for $\pdet$, we are in a scenario of a binomial model with unknown $N_s$ and ``known'' $p$ and thus can estimate $N_s$.
%In these models, point estimates of $N_s$ are unstable unless $\pdet>0.41$ \citep{Olkin1981}.
%One approach to regularizing these estimates is to construct a hierarchical model for site-specific abundance, e.g. $N_s\ind \Po(\lambda)$ \citep{Raftery1988, Royle2004NMixture}. 
%Assuming Poisson abundance, we can decompose $N_{s}$ into observed and unobserved portions: $\nobs \sim Po(\lambda \pdet)$ and, independently, $n_{s}^{(unobs)} \sim \Po\left(\lambda[1-\pdet]\right)$.
%Although alternative distributions could be considered, e.g. negative binomial, our experience with Ovenbird point counts suggests that, after accounting for appropriate explanatory variables, the resulting abundances are likely underdispersed rather than overdispersed, and thus we will use the Poisson assumption here.

@@

Due to the harried process of avian point counts, times to first detection are typically not recorded exactly, but are instead censored into $I$ intervals. 
Let $C_i$ for $i=1,\dots,I$ indicate the right endpoint of the $i$th interval then $C_I$ is the total survey duration and, letting $C_0=0$, the $i$th interval is $(C_{i-1},C_{i}]$. 
Let $n_{si}$ be the number of individuals counted during interval $i$ on survey $s$, $\nobs_s = \sum_{i=1}^I n_{si}$, and $\vn_{s}=(n_{s1},\dots,n_{sI})$.
Assuming independence amongst individuals and sites, we have $\vn_{s} \ind \Mult \left(\nobs_s, \vp_{s}\right)$, where $\vp_{s}=(p_{s1},\dots,p_{sI})$ and is calculated from the TTDD: $p_{si} = F_{T|\dt}(C_i|\dt) - F_{T|\dt}(C_{i-1}|\dt)$.

@@

For ease of reference, the final full model is provided in equation \eqref{eq:model} where the conditioning of the TTDD cdf on $\alpha$ and $\varphi_s$ is made explicit.
\begin{align}
\nobs_s &\ind \Po(\lambda_s \pdet_s) \nonumber\\
\vn_s &\ind \Mult(\nobs_s, \vp_s); \qquad \vp_s = (p_{s1},\dots,p_{sI}) \nonumber\\
\pdet_s &= F_T(C_I|\alpha,\varphi_s) \nonumber \\
p_{s1} &= \left[(1-\gamma) + \gamma\Fm(C_1|\alpha,\varphi_s)\right]/\pdet_s \label{eq:model} \\
p_{si} &= \gamma\left[\Fm(C_i|\alpha,\varphi_s) - \Fm(C_{i-1}|\alpha,\varphi_s)\right]/\pdet_s \nonumber\\
\log(\lambda_s) &= \vX_{s}^A\vbeta^A + \vZ_{s}^A\vxi^A; \qquad \xi_j^A \ind N(0,\sigma_{A[j]}^2) \nonumber\\
\log(\varphi_{s}) &= \vX_{s}^D\vbeta^D + \vZ_{s}^D\vxi^D; \qquad \xi_j^D \ind N(0,\sigma_{D[j]}^2) \nonumber
\end{align}
We adopt a Bayesian approach and therefore require a prior over the model parameters.
To ease construction of a default prior for this model, we standardize all explanatory variables and then construct priors to be diffuse within a reasonable range of values.
Normal prior mean and standard deviation (sd) for the abundance intercept was set at a median abundance of 3 birds per site and a 95\% probability of 0-14 birds present (counted and uncounted).  
Normal prior mean and sd for the detection intercept were chosen so that, based on an intercept-only non-mixture model with $\alpha=1$: (i) median prior detection probability was $p_{s}^{(det)} = 0.50$, and (ii) 95\% of the prior detection probability was within $p_{s}^{(det)} \in (0.01, 1.0)$.  
Normal priors for fixed effect parameters were centered at zero with standard deviations matching the appropriate intercept term.  
All standard deviations and $\alpha$ were given half-Cauchy priors with location 0 and scale 1 for the untruncated Cauchy, and the mixture parameter $\gamma$ was assigned a Unif(0,1) prior in mixture models.
All scalar parameters were assumed independent \emph{a priori}.

We fit the models by MCMC sampling using the Bayesian statistical software Stan, implemented via the R package \texttt{rstan} version 2.8.0 \citep{Rstan2016}.  
%Stan model code and \texttt{rstan} code to run the models can be found in the Supplementary Materials.  
We discarded half of the iterations as warmup and thinned by 10.  
We monitored convergence of the MCMC chains using Geweke z-score diagnostics \citep{Geweke1991} and reran models if lack of convergence was indicated by a non-normal distribution of the z-scores or if the effective sample size for any parameter was below 1000.  
The number of iterations used depended on the model and is detailed later. 
For most models, we accepted Stan defaults for initial values; however, gamma and Weibull models sometimes failed to run unless care was taken in the specification of initial values.

@@

We focus inference on the overall detection probability across all surveys: $\pdet = \sum\limits_{s}\nobs_s \big/$ $\sum\limits_{s}N_{s}$.  
Within simulation studies, we compared model estimates to true $\pdet$ values by percent coverage and by posterior p-values --- the proportion of MCMC samples having a smaller value than the true value.

@@

In the following analyses we distinguish two categories of purely continuous TTDDs: peaked and nonpeaked.  
Detection rates $h(t)$ of peaked distributions generally increase over time, while detection rates of nonpeaked distributions generally decrease over time.
More formally, we define a peaked TTDD as having a mode greater than zero (or $C_1$ for lognormal) while a nonpeaked TTDD has a mode of zero (or less than $C_1$), but we consider exponential TTDDs to be neither peaked nor nonpeaked.

@@

To assess the need for incorporating a mixture component to increase the probability of detection in the initial interval as discussed in Section \ref{s:interval}, we simulated 5600 intercept-only datasets: 100 replicates using 4 values of $\pdet$ (0.50, 0.65, 0.80, and 0.95) from each of 14 TTDDs (each combination of peaked/nonpeaked, mixture/non-mixture, and exponential/gamma/Weibull/lognormal, where exponential models are considered nonpeaked).  
We chose true parameter values (Table \ref{S-simvals}) and the number of surveys (381) to mimic the Ovenbird analysis (Section \ref{sec:ovenbirds}).  
In particular, we set parameters such that (i) in nonpeaked models, 70\% of \textit{detected} individuals were observed during the first two minutes, and (ii) in peaked models, the detection mode for `hard to detect' individuals occured at 5 minutes.

We fit each dataset with two models: mixture and non-mixture versions of the distribution family, e.g. exponential, used to simulate the data.
For each dataset-analysis combination, we sampled $>$ 90,000 iterations which showed no evidence of lack of convergence according to the Geweke diagnostic and reached over 1,000 effective samples for all parameters.

For each dataset and inference model combination, we summarize the analysis across simulations by averaging the posterior median and reporting coverage for 90\% credible intervals.
If analyses are providing reasonable estimates of $\pdet$, we expect the average median to be unbiased and the coverage to be close to 90\%.
Figure \ref{fig:mixture_fig} provides a summary of these quantities. 
When a mixture model is used to simulate the data (lower row of plots), there is clearly a benefit to using a mixture model for inference.  
Using a non-mixture model for inference, the credible interval coverage is near zero for most models with the exponential model overestimating $\pdet$ and the other models underestimating. 
When a non-mixture model is used to simulate the data (upper row of plots), there are no clearly discernable differences between the ability of a non-mixture or mixture model to capture $\pdet$. 
These results support the general default use of a mixture model over a non-mixture model.

For nonpeaked datasets, estimates of $\pdet$ from the same TTDD inference model differed by only 1-5\% between $\pdet=0.50$ and $\pdet=0.65$ simulations (Tables \ref{S-tbl:mix50} to \ref{S-tbl:mix95}), and credible intervals had roughly the same widths.
This suggests that, when data are nonpeaked with true detection probabilities less than 65-80\%, the patterns of detections over time are insufficient for distinguishing between moderate and low values of $\pdet$.
In these cases, mixture inference models estimated higher detection probabilities than did non-mixture models.

\begin{figure}\centering
\includegraphics[width=\textwidth]{"Sims/SimZero/mixture_fig"}
\caption{
Comparisons of mixture and non-mixture TTDDs.  \textit{Left}: Average posterior median detection probabilities across 100 replicate simulations.  Dashed lines show unbiased estimation.   \textit{Right}: Coverage of 90\% credible intervals across 100 replicate simulations.  Dashed lines depict a 95\% range of observed coverage that is consistent with nominal coverage.  \textit{Rows}: Data simulated from non-mixture (upper) or mixture (lower) TTDDs.  \textit{Columns}:  Data simulated from nonpeaked (left), exponential (center), or peaked (right) TTDDs.  Each inference model was fit only to datasets from the same TTDD family (e.g. lognormal to lognormal).
}
\label{fig:mixture_fig} 
\end{figure}
